sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-02 doc/paper.Rmd
Edit

Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs.
GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted.

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed; however, additional options in sdmTMB allow them to be modelled as a random walk or first-order autoregressive, AR(1), process.
We include additional flexibility in specifying the linear fixed effect matrix $\bm{X \beta}$: sdmTMB allows covariates to be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000].

@@

<!-- Paragraph on model selection here-->
Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
Because of the well documented biases of AIC with hierarchical or mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold  cross-validation. 
Our wrapper function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold, and then summarized across folds for a total measure of the predictive accuracy of a model.

<!-- Paragraph on derived quantities here-->
The results from a model fit with sdmTMB can be used to generate quantities of interest, especially relevant to population status.
Estimated parameters from the Matérn covariance function can be used to describe the spatial variance and Matérn range.
We provide additional functionality for generating population-level summaries for each time slice; these metrics include the fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index ( `sdmTMB::get_index()`).

>>>>

<!-- Although sdmTMB has been developed with spatial or spatiotemporal modelling as a primary focus, the statistical models it fits fall under the class of GLMMs. -->
<!-- GLMMs are available in several widely used packages, including glmmTMB [@brooks2017], mgcv [@wood2017a], and lme4 [@bates2015], and to allow comparison with these packages, sdmTMB allows for spatial and spatiotemporal effects to be omitted. -->

By default, we assume that if spatiotemporal fields are included, they are independent and identically distributed (IID); however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \@ref(fig:diagram)).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models; GAMs) using the same `s()` syntax as in mgcv [@wood2017a], or to be modelled with threshold shapes [e.g., hockey stick models, @barrowman2000] (Fig.\ \@ref(fig:diagram)).

@@

<!-- Paragraph on model selection here-->
Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [@akaike1974]. 
Because of well-documented biases of AIC with mixed-effects models [@liang2008] we also provide functionality to perform estimation via k-fold cross-validation. 
Our function `sdmTMB::sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.

<!-- Paragraph on derived quantities here-->
The results from a model fit with sdmTMB can be used to generate quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including fraction of area occupied (TODO?), 2-dimensional center of gravity (`sdmTMB::get_cog()`), and total population index (`sdmTMB::get_index()`).
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range (distance at which spatial correlation decays to $\sim$ 0.13).
